Proving linearizability using forward simulations\thanks{An extended version of this paper including the missing proofs can be found at \cite{extended}.}

### 2017-05-05 draftCAV/stack.tex
Updated AbsS figure.

\begin{figure} [t]
%\vspace{-2mm}
\hspace{-15mm}
\begin{minipage}[t]{7.5cm}
\scriptsize{
{	%\centering
	\begin{align*}
%		& \mathsf{Operations}, \mathsf{Values} : \mathsf{Sort} \\
		& \text{\underline{Shared State}} \\
		& \text{{\ttfamily \bfseries function}}\ \mathsf{present} : \<Ops> \rightarrow \{{\tt true},{\tt false}\} \\
		& \text{{\ttfamily \bfseries function}}\ \mathsf{pending} : \<Ops> \rightarrow \{{\tt true},{\tt false}\} \\
		& \text{{\ttfamily \bfseries function}}\ \mathsf{arg} : \<Ops> \rightarrow \<Vals> \\
		& \text{{\ttfamily \bfseries function}}\ \mathsf{before} : \<Ops> \times \<Ops> \rightarrow \{{\tt true},{\tt false}\} \\
		& \text{{\ttfamily \bfseries function}}\ \mathsf{maxAtInvoc} : \<Ops> \rightarrow 2^{\<Ops>} \\
		& \text{{\ttfamily \bfseries function}}\ \mathsf{overlap} : \<Ops> \rightarrow 2^{\<Ops>} \\
%		& \mathsf{stored}, \mathsf{pending} : \mathsf{Operations} &\mbox{ // the status of an enqueue} \\
%		& \mathsf{arg} : \mathsf{Operations} \rightarrow \mathsf{Values} &\mbox{ // the argument of an enqueue} \\
	\end{align*}
}}
\end{minipage}
\begin{minipage}[t]{7.3cm}
\begin{lstlisting}
void push($\<Vals>$ x) {
  atomic rule inv(push,x,k = new);
  atomic rule ret(push,k);
}

$\<Vals>$ pop() {
  atomic rule inv(pop,k = new);
  atomic rule com(pop,y,k);
  atomic rule ret(pop,y,k);
}
\end{lstlisting}
\end{minipage}

\vspace{-5mm}
\begin{minipage}[t]{6.2cm}
\begin{lstlisting}
rule inv(push,x,k) = {
  $\mathsf{present}$(k) = true;
  $\mathsf{pending}$(k) = true;
  $\mathsf{arg}$(k) = x;
  forall k1 with $\mathsf{present}$(k1)$\land \neg \mathsf{pending}$(k1) do
    $\mathsf{before}$(k1,k) = true;
  forall k1 do
    $\mathsf{overlap}$(k1) = $\mathsf{overlap}$(k1) $\cup$ $\{$k$\}$;
}
rule ret(push,k) = {
  $\mathsf{pending}$(k) = false;
  return;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{5cm}
%\vspace{-3mm}
\begin{lstlisting}
rule inv(pop,k) = {
  forall k1 with $\mathsf{present}$(k1)$\land \neg \mathsf{pending}$(k1)$\land$
                 $\neg \exists k'.\ \mathsf{present}(k')\land \neg \mathsf{pending}(k')\,\land$ 
                 $\mathsf{before}$(k1,$k'$) do
    $\mathsf{maxAtInvoc}$(k) = $\mathsf{maxAtInvoc}$(k) $\cup$ $\{$k1$\}$;
  forall k1 with $\mathsf{present}$(k1)$\land \mathsf{pending}$(k1) do
    $\mathsf{overlap}$(k) = $\mathsf{overlap}$(k) $\cup$ $\{$k1$\}$;
}
rule ret(pop,y,k) = {
  return y;
}
\end{lstlisting}
\end{minipage}

\vspace{-4mm}
\begin{lstlisting}
rule com(pop,y,k) = {
  choose a $\in$ {true, false};
  if ((a $\vee$ $\forall k'.\ \mathsf{present}(k')$=false) $\land$ $\mathsf{maxAtInvoc}$(k)=$\,\emptyset$) 
    y = EMPTY;
  else 
    choose k1 $\in\<Ops>$ with $\mathsf{present}$(k1)$\land$ k1$\in\mathsf{maxAtInvoc}$(k)$\cup\,\mathsf{overlap}$(k);
    $\mathsf{present}$(k1) = false;
    y = $\mathsf{arg}$(k1);
    forall k2 with k1$\in\mathsf{maxAtInvoc}$(k2) do
      $\mathsf{maxAtInvoc}$(k2) = $\mathsf{maxAtInvoc}$(k2)$\setminus\{$k1$\}$;
      forall k3 with $\mathsf{before_1}$(k3,k1)$\land \forall k'.\ (\mathsf{before_1}($k3$,k')\land k'\neq $k1$)\Rightarrow k'\in \mathsf{overlap}($k2$)$ do
        $\mathsf{maxAtInvoc}$(k2) = $\mathsf{maxAtInvoc}$(k2)$\cup\{$k3$\}$;
}
\end{lstlisting}
 %\neg\exists k'.\ \mathsf{present}(k')\,=\,$true$\,\land\, \mathsf{before}($k3,$k')\land \mathsf{before}(k',$k1$)$

%{\scriptsize
%  \centering
%  \begin{mathpar}
%    \inferrule[call-push]{
%      k\not\in dom(cp) \\ 
%      d\neq {\tt EMPTY} \\
%      \forall k'.\ ov'(k') = ov(k')\cup \{k\}
%    }{
%      O,<,\ell,rv,cp,be,ov 
%      \xrightarrow{inv(push,d,k)} 
%      O\cup\{k\},<\cup\ {\tt COMP}(O)\times\{k\},\ell[k\mapsto (d,{\tt PEND})],rv,cp[k\mapsto A_1],be,ov'
%    }\hspace{5mm}
%
%    \vspace{-1mm}
%    \inferrule[call-pop]{
%      k\not\in dom(cp) %\\ k\not\in dom(cp) \\ k\not\in dom(cp)
%    }{
%      O,<,\ell,rv,cp,be,ov
%      \xrightarrow{inv(pop,k)} 
%      O,<,\ell,rv,cp[k\mapsto R_1],be[k\mapsto maxCo(O)],ov[k\mapsto {\tt PEND}(O)]
%    }\hspace{5mm}
%    
%    \vspace{-1mm}
%    \inferrule[ret-pop]{
%       cp(k) = R_2 \\
%       rv(k)=d  
%    }{
%      O,<,\ell,rv,cp,be,ov
%      \xrightarrow{ret(pop,d,k)}
%      O,<,\ell,rv,cp[k\mapsto R_3],be,ov
%    }\hspace{5mm}
%
%    \vspace{-1mm}
%    \inferrule[ret-push1]{
%      cp(k) = A_1 \\
%      k \in O \\
%      \ell(k) = (d,{\tt PEND}) 
%    }{
%      O,<,\ell,rv,cp,be,ov
%      \xrightarrow{ret(push,k)}
%      O,<,\ell[k\mapsto (d,{\tt COMP})],rv,cp[k\mapsto A_2],be,ov
%    }\hspace{5mm}
%    \inferrule[ret-push2]{
%      cp(k) = A_1 \\
%      k \not\in O 
%    }{
%      O,<,\ell,rv,cp,be,ov
%      \xrightarrow{ret(push,k)}
%      O,<,\ell,rv,cp[k\mapsto A_2],be,ov
%    }\hspace{5mm}
%
%    \inferrule[com-pop1]{
%       cp(k) = R_1 \\
%       d\neq{\tt EMPTY} \\
%       k'\in be(k)\cup ov(k) \\
%        \ell_1(k')=d \\
%       \forall k_1.\ k'\not\in be(k_1) \Rightarrow be'(k_1)=be(k_1) \\     \\
%       \forall k_1.\ k'\in be(k_1) \Rightarrow be'(k_1)=(be(k_1)\setminus\{k'\})\cup \{k_2: k_2\in pred_{<}(k') \land \forall k_3. (k_2\in pred_{<}(k_3) \land k_3\neq k') => k_3\in ov(k_1)\} \\
%    }{
%      O,<,\ell,rv,cp,be,ov
%      \xrightarrow{com(pop,d,k)} \\
%      O\setminus \{k'\},<\uparrow k',\ell,rv[k\mapsto d],cp[k\mapsto R_2],be',ov %be[\forall k''.\ k'\in be(k'') \Rightarrow k'' \mapsto (be(k'')\cup pred_{<}(k'))\setminus\{k'\}]
%    }\hspace{5mm}
%
%    %\vspace{-1mm}
%    \inferrule[com-pop2]{
%       cp(k) = R_1 \\
%       be(k)=\emptyset
%    }{
%      O,<,\ell,rv,cp,be,ov
%      \xrightarrow{com(pop,{\tt EMPTY},k)}
%      O,<,\ell,rv[k\mapsto {\tt EMPTY}],cp[k\mapsto R_2],be,ov
%    }\hspace{5mm}
%
%    
%      \end{mathpar}
%  }
 \vspace{-4mm}
  \caption{The abstract stack implementation $AbsS$. We use $\mathsf{before_1}$ to denote the immediate successor relation induced by $\mathsf{before}$, i.e., $\mathsf{before}_1(k,k')={\tt true}$ iff $\neg\exists k''.\ \mathsf{present}(k'')\land \mathsf{before}(k,k'')\land \mathsf{before}(k'',k')$.
%  The transition relation of $AbsQ$. We use the following notions: $maxCo(O)$ is the set of greatest operations in $O$ (w.r.t. $<$) which are completed, i.e., $maxCo(O)=\set{k\in O: \ell_2(k)={\tt COMP}, \forall k'\in O.\ k' < k \vee \ell_2(k')={\tt PEND}}$, ${\tt PEND}(O)=\{k\in O: \ell_2(k)={\tt PEND}\}$, and $pred_{<}(k')$ is the set of immediate predecessors of $k'$ according to $<$, i.e., $pred_{<}(k')=\set{k\in O: k < k'\land \forall k''\in O.\ k'' > k' \vee k'' < k}$.%\textcolor{red}{call-push should have $d \neq EMPTY$ in the premise. If I understand correctly, there is a problem change in $be$ in lin-pop1 rule. What I understand is we move $k''$ to every predecesoor of $k'$. But we should not move it to predecessor of $k'$ if this predecessor has a successor $m$ such that $m \in be(k'')$. Is this case reflected in your definition. Please read sub-bullet 4 of bullet 4 in the definition of $L_I$ in my document. Lin-pop2 seems ok for me. Last note: Bibliography seems not correct to me.}  $f[\forall k.\ k\mapsto expr(k)]$ is the function $g$ defined by $g(k)=expr(k)$,   $g(k)=f(k)$ for all $k$ such that $guard(k)$ is false, and 
  }
  \label{fig:transitions:AbsS}
\vspace{-5mm}
\end{figure}

>>>>

\begin{figure}[t]
  \begin{minipage}[t]{0.39\linewidth}
    \begin{lstlisting}
function loc: $\mathbb{O} \to \set{\tt inv, com, ret, \bot}$
function arg, ret: $\mathbb{O} \to \mathbb{V}$
function present, pending: $\mathbb{O} \to \mathbb{B}$
function before: $\mathbb{O} \times \mathbb{O} \to \mathbb{B}$
function maxAtInvoc: $\mathbb{O} \to \pow{\mathbb{O}}$
function overlap: $\mathbb{O} \to \pow{\mathbb{O}}$

rule inv(push,x,k):
  present(k) := true
  pending(k) := true
  arg(k) := x
  forall k1 with present(k1):
    if $\lnot$pending(k1):
      before(k1,k) := true
  forall k1 with pending(k1):
    overlap(k1) := overlap(k1)$\cup$$\{$k$\}$

rule ret(push,k):
  pending(k) := false

rule ret(pop,y,k):
  assert ret(k) = y
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.59\linewidth}
    \begin{lstlisting}
rule inv(pop,k):
  forall k1 with present(k1):
    if pending(k1):
      overlap(k) := overlap(k) $\cup$ $\{$k1$\}$
    else:
      if $\forall$k2.present(k2)$\land$before(k1,k2)$\Rightarrow$pending(k2):
        maxAtInvoc(k) = maxAtInvoc(k) $\cup$ $\{$k1$\}$

rule com(pop,y,k):
  ret(k) := y
  if y = EMPTY:
    assert maxAtInvoc(k) = $\emptyset$
  else:
    let k1 := arg$^{-1}$(y)
    assert present(k1)
    assert k1 $\in$ maxAtInvoc(k) $\cup$ overlap(k)
    present(k1) := false
    forall k2 with k1 $\in$ maxAtInvoc(k2):
      maxAtInvoc(k2) := maxAtInvoc(k2) $\setminus$ $\{$k1$\}$
      forall k3 with before$_1$(k3,k1):
        if $\forall$k4$\neq$k1.before$_1$(k3,k4) $\Rightarrow$ k4$\in$overlap(k2):
          maxAtInvoc(k2) := maxAtInvoc(k2) $\cup$ $\{$k3$\}$
    \end{lstlisting}
  \end{minipage}
  \caption{The $AbsS$ implementation; each rule \lstinline|$\alpha$(\_,k)|
  implicitly begins with \lstinline|assert loc(k)=$\alpha$| and ends with the
  appropriate \lstinline|loc(k):=$\beta$|; and \lstinline|before$_1$| denotes
  the transitive reduction of \lstinline|before|.}
  \label{fig:transitions:AbsS}
\end{figure}
